---
title: "blockcpd: detecting multiple change-points in multiple signals"
#author: 
#- name: Lucas Prates 
#- name: Florencia Leonardi
output: github_document
#rmarkdown::html_vignette
#vignette: >
#  %\VignetteIndexEntry{blockcpd_simulation_vign}
#  %\VignetteEngine{knitr::rmarkdown}
#  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
```

  Welcome to `blockcpd`! This `R` package allows the user to fit change-point 
  detection models for the task of segmenting repeated signals. It also provides
  tools for plotting, selecting and comparing the fitted models.
 
  The package tries to be ease of use, yet providing some flexibility. The user 
  can choose the fitting method, statistical family, basic constraints and 
  define his own penalization function.
  
  For an in-depth mathematical formulation of the model and its properties, see
  the reference on the end. Summarizing, it performs an offline multiple 
  change-point segmentation based on regularized likelihood. The signals are 
  assumed be independent and identically distributed, and have the same size. We 
  are interested in extending the applicability in the future.
 
  An interesting functionality the user might want to try is the confidence
  plot: a graphic that shows the estimates of how likely it is for a structural 
  break to occur at a given index.
  
  We now provide a quick tutorial on how to install the package, simulate 
  data, fit the model, avoid over-fitting, and exploring the confidence plot
  feature.
  
## Installation

  To install the package from this repository, install the
[devtools](https://cran.r-project.org/web/packages/devtools/index.html) package 
and run the command below.
  
```{r eval = FALSE}
devtools::install_github("https://github.com/Lucas-Prates/blockcpd")
```

## Examples
### Simulating data 
  
  Lets simulate data with $20$ signals (samples), each with variables, segmented 
  in four intervals: $[1, 50]$, $[51, 110]$, $[111, 180]$ and $[181, 200]$. We 
  consider that the variables from each block comes from a 
  [exponential distribution](https://en.wikipedia.org/wiki/Exponential_distribution) 
  with scale parameters $1$, $2$, $3$ and $1$ respectively. 
  
  To simulate from this model, we use the `rcpd` function provided by the
  package. To pass the parameters, provide a list, where the keys are
  the parameter name for the model, and the value is a vector whose entries
  are the value of that parameter for each block. The blocks are specified
  with the `changepoints` argument, which is a vector containing the end point
  of each block except the last, in our case $(50, 110, 180)$. 
  
  We also must specify the statistical family. Currently, the 
  package implements the Normal (Gaussian)  with unknown mean and variance, 
  Poisson, Exponential, Bernoulli and Binary Markov Chain (2-states) 
  distributions.
   
```{r}
library(blockcpd)
set.seed(204)
parameters = list(scale = c(1, 2, 3, 1))

# nrow = number of signals
# ncol = number of variables or observations per signal
sim_df <- rcpd(nrow = 20, ncol = 200, 
               family = "exponential", parameters = parameters,
               changepoints = c(50, 110, 180)) 
```

  The output is a list with three elements: the data matrix, the true
  change-points and parameters.
  
  The figure below show how the first four signals look like. The vertical
  dashed lines mark the location of the change-points.
  
```{r, echo=FALSE, fig.width=8, fig.height = 7}
par(mfrow =  c(2, 2))
plot(sim_df$data_matrix[1, ], ylab = "Signal", main = "Signal 1")
abline(v = sim_df$changepoints, col = 'red', lty = 'dashed')
plot(sim_df$data_matrix[2, ], ylab = "Signal", main = "Signal 2")
abline(v = sim_df$changepoints, col = 'red', lty = 'dashed')
plot(sim_df$data_matrix[3, ], ylab = "Signal", main = "Signal 3")
abline(v = sim_df$changepoints, col = 'red', lty = 'dashed')
plot(sim_df$data_matrix[4, ], ylab = "Signal", main = "Signal 4")
abline(v = sim_df$changepoints, col = 'red', lty = 'dashed')
par(mfrow =  c(1, 1))
```

### Fitting the model 

  Now that we have a data set, we fit the model using the `fit_blockcpd`
  function. The user can choose fitting method, statistical family, maximum 
  number of blocks and provide his own penalization function.
  
  Two fitting methods are currently implemented: "hierseg" and "dynseg". The
  first is a fast greedy implementation of the hierarchical segmentation (also
  called binary segmentation), and the second is dynamic segmentation, an exact
  but slower implementation using dynamic programming. The statistical methods
  are the same as discussed in the simulations. 
  
  We will better discuss the penalization constant and maximum blocks in the 
  next section.
  
  Lets fit the model using "hierseg" (default), the "exponential" family, 
  penalization constant equals 1 and default default penalization.

```{r}
  seg_model = fit_blockcpd(sim_df$data_matrix, family = "exponential", 
                           lambda = 1)
```

  The return object is a S3 class called `blockcpd`. It contains the estimated
  change-points, parameters, loss, negative log likelihood and metadata. It also
  contains bootstrap information if the bootstrap flag is `TRUE`. We will see
  why running the model with bootstrap can be useful for the practitioner.
  
  The user can plot how a parameter varies with the indices just by calling 
  `plot` and passing the model as an argument. 
  
```{r, fig.width=8, fig.height = 4}
  plot(seg_model, parameter = "scale")
```
 
  The flat regions corresponds to variables grouped in the same block. The
  height of the block corresponds to the estimated value of the parameter for
  that block. The vertical lines shows where the model detected a change-point.
  In this example, it detected the changes at $(50, 110, 180)$, which 
  are the also true change-points.
  
  Since the scale parameter is associated with the expected value of the 
  variables for the exponential distribution, we could plot a "average signal"
  along with the block plot, as show in the figure below.
  
```{r, echo=FALSE, fig.width=8, fig.height = 4}
  mean_signal = apply(sim_df$data_matrix, 2, mean)
  blockcpd_obj = seg_model
  ncp = blockcpd_obj$ncp
  changepoints = blockcpd_obj$changepoints
  parameter_vec = blockcpd_obj$parameters[["scale"]]
  sf = stepfun((1:200)[changepoints], parameter_vec)
  plot(sf, xlim = c(1, 200), do.points = F, xaxs = "i",
       xlab = "Index", ylab = "Scale (Average)",
       ylim = c(0, max(mean_signal) + 1),
       main = paste("Block plot along with 'average signal'"))
  points(mean_signal, col = 'blue')
```

  The black lines are the estimated scale (average value) parameter for 
  the blocks, and the blue points are the average signal for each index.
  
### Avoiding over-segmentation

  In order to avoid over segmentation, two parameters can be used: the
  penalization constant (the `lambda` argument in `fit_blockcpd`) and the 
  maximum number of blocks (the `max_blocks` argument).
  
  If we used $0.1$ as the penalization constant, we would have detected $86$ 
  blocks, a clear over-segmentation! However, we only know it would be a
  over-segmentation because we simulated the data. We need an data-driven 
  approach to select lambda. 
  
  The constant has to be chosen carefully. A common approach is to use an 
  adaptation of the [elbow plot](https://en.wikipedia.org/wiki/Elbow_method_(clustering)) 
  heuristics. A graphic of how the number of detect blocks varies with
  the penalization constant lambda is constructed, and then we try to visually
  select a value in which the curve "flats out" after it.
  
  The package provides the `elbow_plot` function which constructs the plot given
  the data set and model arguments. The model arguments must be passed as a 
  list, and the constant value should *not* be included.
  
```{r, fig.width=8, fig.height = 4}
  model_args = list(family = "exponential") # do not include lambda!

  # search space for lambda
  lambda_left = 0
  lambda_right = 5
  step = 0.5 # distance between lambda values
  
  # creates the plot and returns output related to the plot 
  # If no return value is passed (no 'ep_info' on the left), then it only
  # plots 
  ep_info = elbow_plot(sim_df$data_matrix, lambda_left = lambda_left,
                       lambda_right = lambda_right, step = step,
                       model_args = model_args)

```
  
  A simple function call plots the result. If the user passes a variable for 
  assignment, the function returns additional information. The function also
  makes an "educated guess" on the value of the constant, in this case $1.5$, 
  justifying the value we used on the beginning.
  **Do not restrict yourself to the suggestion, specially for very small values of step.**
  
  Another way to control over-segmentation is using the `max_blocks`
  argument. When working with a large number of variables, if there is no 
  justification for a your signal to have a large number of blocks, try setting
  small values for `max_blocks`. This will also make `fit_blockcpd` run much
  faster.
  
### Is that change spurious? Check it with Confidence Plot

  A central question is Statistics is: how confident are you on the fitted
  model? How do you know the results you obtained are not spurious, that is,
  they are not due to chance?
 
  If you are a practitioner and your experience suggests that your signals have 
  a structural break at given location, it is not enough to fit a model and 
  detect a change at the location; you have to argue that, if you repeated the 
  experiment, you would consistently detect a change at that location. This is 
  the core of scientific inquiry.
  
  To that end, we introduce the confidence plot. It plots the estimates of how 
  likely it is for the model to detect a change at any given point. True 
  structural breaks should have confidence near 100%, while non-break indices 
  should have a confidence near 0%. It might also be difficult to detect a 
  true change-point at the given sample size. In this case, it should fluctuate 
  in the middle.
  
  The confidence is computed using bootstrap re-sampling. To create the plot,
  we first need to refit the model passing the `TRUE` value to the `bootstrap` 
  argument. Then, we call the `confidence_plot` passing the fitted model.
 
```{r, fig.width=8, fig.height = 4}
  seg_model = fit_blockcpd(sim_df$data_matrix, family = "exponential", 
                           lambda = 1, 
                           bootstrap = TRUE, bootstrap_samples = 200L)
  confidence_plot(seg_model, scale = "percentage")
```
  
  The plot shows the detection percentage of each index as a change-point. The
  dashed vertical red lines shows the location of the final detected 
  change-points. 
  
  In this example, it is strongly suggested that $50$ and $180$
  are true change-points, but $110$ also has a high detection value near $70$%.
  Notice that the index $1$ was detected around $25$% of the time, which is due
  to chance. If we collected more signals, this value would approach $0$%.

### Recommendation for datasets with large number of variables
  
  To reduce run-time on large datasets, we recommend using the `hierseg` method
  when using `fit_blockcpd` and `elbow_plot`. It is also advisable to set small
  values for `max_blocks`, not only for time issues but also to avoid 
  over-segmentation.
  
  A final observation is that fitting the model with the `bootstrap` as `TRUE` 
  is computationally expensive, so avoid using very large values for 
  `bootstrap_samples`.
  
### Reference

  This package is an implementation of the method and heuristics discussed in 
  the [paper](https://arxiv.org/abs/2111.10187) submitted to the Journal of 
  Applied Statistics, currently available as a pre-print. For citation purposes, 
  see the reference below.
  
  Prates, L., Lemes, R. B., Hünemeier, T., & Leonardi, F. (2021). Population based 
  change-point detection for the identification of homozygosity islands. arXiv 
  preprint arXiv:2111.10187.

